Layered Grammar for Alluvial Plots

### 2020-01-17 paper.rmd
revise & extend comparisons with other packages

[The titular vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the package documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for comprehensive examples. In brief, the package contains stat and geom functions to add the following layers to a ggplot2 object:

@@

While questions and requests by users have prompted many improvements, the current version is feature complete with one exception: improvements and alternatives to the flow elements connecting lodes. The current splines will remain the default.

@@

The diagrams produced by ggalluvial are statistical graphics in the sense that they communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006]. The package adopts for them the term _alluvial plot_.[^plots]
These plots are distinct from Sankey diagrams (they do not necessarily represent flow) and constitute a subtype of parallel sets plots distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them, features shared only by ggparallel.
The package includes several levers of control over the statistical transformations, and thereby over the messages conveyed by the plot.[^easy]
The plot elements are rendered by separate layers, following the additive (`+`) syntax of ggplot2, a feature shared only by ggpcp.

[^plots]: This has the unfortunate side effect of conflating search results from the geology literature.
[^easy]: Other ggplot2 extensions are not so flexible; in particular, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for more expedient data exploration.

Sankey, parallel sets, and alluvial diagrams are often conflated, and there is currently no consensus on what features are distinctive to each type.
Moreover, their graphical elements go by a variety of names, often interchangeably.
While adopting new terms risks adding to this confusion, these alluvial plots are, i think, distinctive enough to warrant it.

>>>>

[The primary vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the package documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for comprehensive examples. In brief, the package contains stat and geom functions to add the following layers to a ggplot2 object:

@@

@@

Sankey, parallel sets, and alluvial diagrams are often conflated, and there is currently no consensus on what features are distinctive to each type.
Moreover, their graphical elements go by a variety of names, often interchangeably.
In order to more clearly describe the features of ggalluvial in relation to similar packages, it is useful to adopt a formal demarcation among these diagram types.

_Statistical graphics_ (here also simply called "plots") are diagrams that communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
By design, graphics produced by ggplot2 extensions (and by the alluvial package) are plots: The stat, geom, and other layers of a ggplot object exactly reproduce a graphic from data (under the same parameter settings).[^exceptions]
Sankey diagrams are much more flexible: The earliest engine efficiency diagrams in this tradition could take a variety of forms to depict the same energy flow and were differently annotated for different audiences [@Schmidt2008], and software implementations may be interactive and enable manual adjustments of graphical elements [@Riehmann2005].
Moreover, the plots generated by these extensions are suitable for a wide variety of data types, whereas Sankey diagrams are still overwhelmingly used to represent flow. Arguably, the extensions are better understood as belonging to a different type of diagram.

[^exceptions]: There are exceptions. For example, the jitter geom in ggplot2 introduces randomness to symbol positions, and the repel geoms of ggrepel [@Slowikowski2019] use heuristic algorithms to position text.

Parallel sets plots might be viewed as a special subtype of Sankey diagram. In turn, the plots produced by the aforecited packages are all parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]
They vary in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
The geoms of ggalluvial offer only a one-dimensional continuum between straight lines and x-splines for the ribbon shapes, though alternatives are planned.
The stats vertically arrange the elements (boxes and ribbons) at each axis. ggalluvial provides more levers of control over these statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]
Notably, in ggalluvial the plot elements are rendered by separate layers, following the additive (`+`) syntax of ggplot2, a feature shared by ggforce and ggpcp but not by ggparallel.

[^easy]: Other ggplot2 extensions are not so flexible; in particular, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for more expedient data exploration.

The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces,[^plots] which are distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them. The first feature is shared by the other packages but are not essential to parallel sets plots; such plots could arrange boxes corresponding to repeated categorical decompositions differently at different axes. While most of the packages separate boxes at each axis with gaps, these can be reduced to zero, so that each package can create alluvial plots. (ggparallel and ggalluvial alone _only_ produce alluvial plots.)
This feature is particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
It is for this reason that i believe the new geological terminology is warranted.

[^hammock]: The possible exceptions are the hammock plots and common angle plots of ggparallel, which are contrasted with a stricter definition of parallel sets plots than i use here, in which ribbons are straight, their widths aggregate to box widths, and they meet without overlap at the sides of boxes, partitioning them [@Hofmann2013].
[^plots]: This has the unfortunate side effect of conflating search results from the geology literature.
